Title Goes Here

### 2023-01-07 inst/paper/_conclusions.qmd
incorporate feedback from coauthors

This preliminary study outlines the benefits of using calculus to target a
variety of compounds that could have been consumed as medicine or diet. 
This method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in historic
documentation, such as lower status, children and women.
We also
show that there are many limitations that will need to be addressed going forward
with this type of analysis, and stress the need for more systematic research on
the consumption of alkaloid-containing items and their subsequent concentration
and preservation in dental calculus,
in addition to how mode of consumption may affect concentrations on different parts
of the dentition.
Another limitation of dental calculus as a medium is the inter- and intra-individual
variability of its formation and the many factors that can influence incorporation
and retention of molecules and particles; however, in the absence
of hair and serum (quite uncommon in archaeology), dental calculus represents
an impressive long-term reservoir of information regarding the consumption of various
alkaloids, whether dietary, medicinal, recreational, or otherwise.

>>>>

This preliminary study outlines the benefits of using calculus to target a
variety of compounds that could have been consumed as medicine or diet. 
This method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in historic
documentation, such as rural communities, children and women.
We also
show that there are many limitations that will need to be addressed going forward
with this type of analysis, and stress the need for more systematic research on
the consumption of alkaloid-containing items and their subsequent concentration
and preservation in dental calculus,
in addition to how mode of consumption may affect concentrations on different parts
of the dentition.
Another limitation of dental calculus as a medium is the inter- and intra-individual
variability of its formation and the many factors that can influence incorporation
and retention of molecules and particles; however, in the absence
of hair and serum (quite uncommon in archaeology), dental calculus represents
an impressive long-term reservoir of information regarding the consumption of various
alkaloids, whether dietary, medicinal, recreational, or otherwise.

### 2023-01-07 inst/paper/_discussion.qmd
incorporate feedback from coauthors

<!-- Big summary/important statement for the science -->
In this study we were able to extract and identify multiple alkaloids from the
dental calculus of individuals from Middenbeemster, a 19th century Dutch
archaeological site. We applied ultra-high-performance liquid chromatography-tandem
mass spectrometry (UHPLC-MS/MS), using a method that was validated by co-occurrence
of drugs and metabolites in dental calculus and blood [@sorensenDrugsCalculus2021].
Here we have shown that the method can also be successfully applied to archaeological
dental calculus.
We extend findings from previous studies on alkaloids in archaeological samples
by extracting multiple different alkaloids from dental calculus, including
cannabinoids, cocaine, nicotine, cotinine, caffeine, theophylline, and salicylic
acid in multiple individuals. Of these compounds, the cannabinoids and cocaine
could not be replicated in a second run on different samples from the same individuals.
We discuss the implications of these findings in light of historical and
archaeological evidence for the consumption of these drugs.

@@

<!-- nicotine and cotinine: tobacco, obviously...-->
Nicotine and its principal metabolite, cotinine, were strongly positively correlated,
both in concentration and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
The detection of nicotine and cotinine is not surprising, as tobacco-use and
pipe-smoking in the Beemsterpolder is well-documented in the literature
[@aten400JaarBeemster2012; @boumanBegravenis2017], and visible on the skeletal
remains as pipe notches [@lemmersMiddenbeemster2013].
There is also documented medicinal use of nicotine in the Beemsterpolder,
where a tobacco-smoke enema was used for headaches, respiratory problems, colds,
and drowsiness from around 1780 to 1830 [@aten400JaarBeemster2012].
In our sample, we also detected nicotine and cotinine (replicated)
in an old adult, probable female individual.
It is unlikely that the compounds entered the dental calculus through
pipe-smoking, as the individual had no visible pipe notches.
More likely that the tobacco entered through an alternate mode of consumption,
secondhand smoke, or the aforementioned tobacco-smoke enema.

@@

<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
While a major limitation is the uncertainty surrounding whether or not a compound
is actually absent, the power of the method is in detecting food and other items
that are consumed in a manner that does not allow macroscopic detection. Taking
tobacco consumption as an example; while pipe
notches are a useful way to identify tobacco consumption, pipe smoking was not the
only mode of tobacco consumption, with others including chewing, drinking, cigars,
and snuff [@goodmanTobaccoHistory1994, p.67]. Pipe-smoking
was mainly practised by males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018],
so chromatographic methods, such as the one presented here,
are suitable for exploring tobacco consumption in an entire population, rather
than a trivial subset of past populations.<!--admittedly, this study did just that...-->

>>>>

<!-- Big summary/important statement for the science -->
In this study we were able to extract and identify multiple alkaloids and salicylic
acid from the
dental calculus of individuals from Middenbeemster, a 19th century Dutch
archaeological site. We applied ultra-high-performance liquid chromatography-tandem
mass spectrometry method using pneumatically assisted electrospray ionisation
(UHPLC-ESI-MS/MS), a method that was validated by co-occurrence
of drugs and metabolites in dental calculus and blood [@sorensenDrugsCalculus2021].
Here we have shown that the method can also be successfully applied to archaeological
dental calculus.
We extend findings from previous studies on alkaloids in archaeological samples
by extracting multiple different alkaloids from dental calculus, including
cannabinoids, cocaine, nicotine, cotinine, caffeine, theophylline, and salicylic
acid in multiple individuals. Of these compounds, the cannabinoids and cocaine
could not be replicated in a second run on different samples from the same individuals.
We discuss the implications of these findings in light of historical and
archaeological evidence for the consumption of these drugs.

@@

<!-- nicotine and cotinine: tobacco, obviously...-->
Nicotine and its principal metabolite, cotinine, were strongly positively correlated,
both in concentration and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
The detection of nicotine and cotinine is not surprising, as tobacco-use and
pipe-smoking in the Beemsterpolder is well-documented in the literature
[@aten400JaarBeemster2012; @boumanBegravenis2017], and visible on the skeletal
remains as pipe notches [@lemmersMiddenbeemster2013].
There is also documented medicinal use of nicotine in the Beemsterpolder,
where a tobacco-smoke enema was used for headaches, respiratory problems, colds,
and drowsiness from around 1780 to 1830 [@aten400JaarBeemster2012].
In our sample, we also detected nicotine and cotinine (replicated)
in an old adult, probable female individual.
In this particular case it is unlikely that the compounds entered the dental
calculus through pipe-smoking, as the individual had no visible pipe notches;
more likely the tobacco entered through an alternate mode of consumption,
secondhand smoke, or the aforementioned tobacco-smoke enema.

@@

<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
While a major limitation is the uncertainty surrounding whether or not a compound
is actually absent, the power of the method is in detecting food and other items
that are consumed in a manner that does not allow macroscopic detection. Taking
tobacco consumption as an example; while pipe
notches are a useful way to identify tobacco consumption, pipe smoking was not the
only mode of tobacco consumption, with others including chewing, drinking, cigars,
and snuff [@goodmanTobaccoHistory1994, p.67]. Pipe-smoking
was mainly practised by males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018],
so chromatographic methods, such as the one presented here,
are suitable for exploring tobacco consumption in an entire society, rather
than a trivial subset of past populations.<!--admittedly, this study did just that...-->

### 2023-01-07 inst/paper/_intro.qmd
incorporate feedback from coauthors

<!--how are compounds incorporated in dental calculus-->
Alkaloids may enter the oral cavity via two pathways: (1) direct
incorporation through oral consumption of alkaloid-containing plants, whether
deliberate or accidental;
and (2) passive diffusion as toxins are transferred from serum to saliva
[@coneInterpretationOral2007],
and then into the oral cavity through the salivary glands. The relation to serum
is why there is often a close correlation between presence (not concentration)
of drugs in oral fluid and blood/serum
[@coneInterpretationOral2007; @willeRelationshipOral2009; @milmanOralFluid2011].
The second pathway allows the
detection of compounds that not only are consumed orally, but also any other mode
of consumption or administration<!--cite e.g. injection or topical-->.
<!-- Additionally, alkaloids will spend shorter time in mouth during direct consumption than
during excretion through saliva, so we expect the second pathway to be more
feasible for detection. -->

@@

<!--brief description of method validation-->
The method we use in this study was developed in a previous study on dental
calculus from cadavers and validated by comparing the results to compounds detected
in the blood of the same individuals.
All compounds that were detected in the blood were also detected in dental calculus,
with additional compounds present in dental calculus that were not present
in blood, suggesting that dental calculus represents a comprehensive history of
consumption over a long time <!-- rephrase-->[@sorensenDrugsCalculus2021].
<!--summary of results-->
We were able to detect both parent compounds and metabolites, including caffeine,
nicotine, theophylline, and cotinine, in the dental calculus of individuals from
a 19th century Dutch population from Middenbeemster. By detecting these compounds
we are able to show the consumption
of tea and coffee and smoking of tobacco on an individual scale, which is also
confirmed by historic documentation and identification of pipe notches in the
dentition.

>>>>

<!--how are compounds incorporated in dental calculus-->
Alkaloids may enter the oral cavity via two pathways: (1) direct
incorporation through oral consumption of alkaloid-containing plants, whether
deliberate or accidental;
and (2) passive diffusion as alkaloids and other compounds are transferred from
serum to saliva, and then into the oral cavity through the salivary glands
in the hours-to-days following consumption [@coneInterpretationOral2007]. The
relation to serum
is why there is often a close correlation between presence (not concentration)
of drugs in oral fluid and blood/serum
[@coneInterpretationOral2007; @willeRelationshipOral2009; @milmanOralFluid2011].
The second pathway allows the
detection of compounds that not only are consumed orally, but also any other mode
of consumption or administration, as long as it is excreted through saliva<!--cite e.g. injection or topical The second pathway may also be more
detectable, as the excretion of compounds through saliva may take hours to days
whereas compounds spend less time in the oral cavity during direct consumption-->.

@@

<!--brief description of method validation-->
In this study we use a ultra-high-performance liquid chromatography-tandem mass
spectrometry method using pneumatically assisted electrospray ionisation
(UHPLC-ESI-MS/MS) that was developed in a previous study on dental
calculus from cadavers and validated by comparing the results to compounds detected
in the blood of the same individuals.
All compounds that were detected in the blood were also detected in dental calculus,
with additional compounds present in dental calculus that were not present
in blood, suggesting that dental calculus represents a comprehensive history of
consumption over a long period of time [@sorensenDrugsCalculus2021].
<!--summary of results-->
We were able to detect both parent compounds and metabolites, including caffeine,
nicotine, theophylline, and cotinine, in the dental calculus of individuals from
a 19th century Dutch population from Middenbeemster. By detecting these compounds
we are able to show the consumption
of tea and coffee and smoking of tobacco on an individual scale, which is also
confirmed by historic documentation and identification of pipe notches in the
dentition.

### 2023-01-07 inst/paper/_materials.qmd
incorporate feedback from coauthors

To reduce the number of factors to account for
in the analysis, we preferentially selected males from the middle adult age category
(35-49). The sample consists of `r nrow(filter(demography, sex == "m"))` males,
`r nrow(filter(demography, sex == "pm"))` probable males,
`r nrow(filter(demography, sex == "pf"))` probable female, and
`r nrow(filter(demography, sex == "f"))` female.
We selected males due to a higher
occurrence of dental calculus deposits than females (unpublished observation).
Dental calculus was preferentially selected from incisors.

@@

# create tables for demographics if there are tables to spare in the article
first_batch <- demography %>%
  #left_join(select(metadata, id, replicated)) %>%
  mutate(sex = factor(sex, levels = c("f", "pf", "pm", "m"))) %>% 
  ggplot(aes(x = age, fill = sex)) +
    geom_bar() +
    scale_fill_viridis_d(drop = FALSE, labels = c("Female", "Female?", "Male?", "Male")) +
    theme_minimal() +
    labs(y = "Count", fill = "Sex") +
    theme(
      panel.grid.major.x = element_blank(),
      axis.title.x = element_blank()
    )

@@

grid::grid.draw(grid::textGrob("Age-at-death", x =.45, y=0.01))

# demography %>% 
#   count(sex) %>%
#   knitr::kable()
# 
# demography %>%
#   count(age) %>% 
#   knitr::kable()
```

>>>>

To reduce the number of factors to account for
in the analysis, we preferentially selected males from the middle adult age category
(35-49). The sample consists of `r nrow(filter(demography, sex == "m"))` males,
`r nrow(filter(demography, sex == "pm"))` probable males,
`r nrow(filter(demography, sex == "pf"))` probable females, and
`r nrow(filter(demography, sex == "f"))` female [@fig-sample-demography].
We selected males due to a higher
occurrence of dental calculus deposits than females (unpublished observation).

@@

# create tables for demographics if there are tables to spare in the article
first_batch <- demography %>%
  #left_join(select(metadata, id, replicated)) %>%
  mutate(sex = factor(sex, levels = c("f", "pf", "pm", "m"))) %>% 
  ggplot(aes(x = age, fill = sex)) +
    geom_bar() +
    scale_fill_viridis_d(drop = FALSE, labels = c("Female", "Female?", "Male?", "Male")) +
    theme_minimal() +
    labs(y = "Count", fill = "Sex") +
    theme(
      panel.grid.major.x = element_blank(),
      axis.title.x = element_blank(),
      plot.margin = unit(c(0,0,0.5,0), "cm")
    )

@@

grid::grid.draw(grid::textGrob("Age-at-death", x =.45, y=0.045))
```

### 2023-01-07 inst/paper/_methods.qmd
incorporate feedback from coauthors

New dental calculus samples were taken under sterile conditions in a positive
pressure laminar flow hood in a dedicated dental calculus lab at Leiden University.
The surface of the tooth was lightly brushed with a sterile, disposable toothbrush
to get rid of surface contaminants. A sterile dental curette was then used to
scrape calculus from the tooth onto weighing paper, which was transferred to 1.5 ml
Eppendorf tubes.
Where possible, previously sampled material was used to prevent repeated sampling
of individuals. The sampling methods can be found in the published material
[@ziesemer16SChallenges2015; @ziesemerGenomeCalculus2018]. 
The samples were sent to the Department of Forensic Medicine at Aarhus University
for ultra-high-performance liquid chromatography tandem mass spectrometry (UHPLC-MS/MS).

The UHPLC-MS/MS method was validated in a separate study on cadavers received
for forensic autopsy and toxicological analysis. Results from dental calculus were
validated against compounds detected in whole blood samples from the same
individuals.

Briefly, dental calculus was treated with citric acid and the dissolution extracts
were cleaned using weak and strong polymeric cation-exchange sorbents.

[[ UHPLC method description needs to be elaborated ]]
<!--rephrase and expand-->

<!-- the samples were washed in 0.5 mL MeOH for 10 seconds, and centrifuged
at 10,000 $\times$ g for 1 minute. This process was repeated. Supernatants were
mixed with 10 mL dimethyl sulphoxide (DMSO) and 50 mL SIL-IS solution in a conical
glass centrifuge tube. The solvent was evaporated under a stream of nitrogen at
40&deg;C until ca. 10 mL residue remained, which was vortexed with 50 ml 30% (v/v)
MeOH. The calculus residue was air-dried in the dark for 24 h and weighed. The
washing/drying procedure was repeated twice using 1 mL MeOH, resulting in three
wash extracts for analysis. -->
In the original method, samples were washed three times to remove residual
substances from the surface of the calculus that originated from oral fluids,
and only extract substances from the calculus. In our samples the washes served
to remove potential contaminants from the burial environment and post-excavation
handling.
<!-- 
The remaining calculus was dissolved using lysing tube beads in 800 $\mu$L of 0.5
$\small{M}$ citric acid (CA) and 50 $\mu$L SIL-IS solution. During dissolution the
suspension was kept at ambient temperature for 1 h under continuous gentle shaking,
and subsequently mixed with 800 $\mu$L MeOH and centrifuged at 10,000 $\times$ g for
5 mins.
-->
<!--
2.4.3. Clean-up by WCX
The dissolution extract was mixed with 6300 m L of water,
100 m L of a 25% ammonia solution to obtain a pH of 6–7 and 50 m L
of the SIL-IS working solution. The solution was passed through a
WCX SPE cartridge with a maximum flow rate of 0.5 mL/min
(mainly using gravitational force (GF)). The cartridge was pre-
conditioned with 1 mL of MeOH followed by 1 mL of water. The
passed extract was collected in a glass tube for clean-up by SCX.
The cartridge was washed with 1 mL of water, then full vacuum
( 70 kPa) was applied for 5 s to remove the bulk volume of water
with a minimum of air flow passing through the sorbent. The wash
water was collected in the same tube. The substances were eluted
from the column using a 2-mL volume of freshly prepared 2% FA in
MeOH. A 10- m L volume of DMSO was added to the eluate, and then
the solvent was evaporated to a residual volume of approximately
10 m L under a stream of nitrogen at 40  C. The residue was vortexed
with a 50- m L volume of 30% MeOH, and the solution was
transferred to an autosampler vial. This solution was used for
the analysis of most substances.

2.4.4. Clean-up by SCX
The sample fraction saved from WCX clean-up was mixed with
100 m L of 12 M HCl and passed through an SCX SPE cartridge that
was pre-conditioned with 1 mL of MeOH followed by 1 mL of
water. The cartridge was washed with 1 mL of water, then full
vacuum was applied for 5 s before it was eluted with 2 mL of 5%
ammonia in MeOH. The eluate was mixed with 10 m L of DMSO
and evaporated to a residual volume of approximately 10 m L
under a stream of nitrogen at 40  C. The residue was vortexed
with a 50- m L volume of 30% MeOH, and the solution was
transferred to an autosampler vial. This solution was used for the
analysis of morphine-3-glu, morphine-6-glu, pregabalin, gaba-
pentin and salicylic acid.

2.4.5. UHPLC
A 5- m L volume was injected onto the column running 99%
mobile phase A. The column flow rate was 400 m L/min, and the
column temperature was maintained at 40  2  C. The mobile
phase was changed through a linear gradient to 0% A and 100% B
over 5 min. Nine minutes after the initial injection, the gradient
was returned to 99% A over 0.1 min, and the column was
equilibrated for 2.9 min before the next injection, resulting in a
total run time of 12 min. The same column and gradient
programme were used for the wash, WCX and SCX extracts. The
eluent was diverted to waste during the time intervals of 0–1.5 and
8–12 min after injection using a post-column switch.

2.4.6. MS-MS
The probe temperature (TEM) was set to 400  C. The curtain
gas (CUR), ion source gas 1 (GS1), ion source gas 2 (GS2) and
collision gas (CAD) were set at 20, 80, 80 and 8 psi, respectively.
The mass spectrometer was operated in both positive ion mode
and negative ion mode at unit mass resolution using polarity
switching to obtain the maximum sensitivity for both analytes
that favour positive ionisation and analytes that favour negative
ionisation. The ion spray voltages were set at 4.5 kV. Scheduled
multiple reaction monitoring (MRM) with windows of 30 s was
applied, and at least 12 data points were obtained across each
peak. The applied MRM conditions are listed in Supplementary
Table S2. Nitrogen was used as the CAD gas. Data acquisition and
processing were performed using Analyst 1.7 and MultiQuant
3.0.3 (Sciex), respectively. -->

For more details about the method and validation, see the original study by
Sørensen and colleagues [-@sorensenDrugsCalculus2021].

Of the `r nrow(demography)` individuals that were analysed,
`r nrow(filter(metadata, replicated == TRUE))` individuals were subsequently
replicated using leftover calculus from the original DNA study.
in <!-- the same lab? different lab? -->.

@@

<!-- methods for age, sex, and stature estimation -->
Demographic and pathological analyses were conducted in the Laboratory for Human
Osteoarchaeology at Leiden University. Sex was estimated using cranial
and pelvic morphological traits [@Standards1994]. Age was estimated using dental wear,
auricular and pubic surface appearance, cranial suture closure, and epiphyseal fusion 
[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985].
<!-- differential diagnoses -->

@@

Various pathological conditions were included in the analysis and were
dichotomised to presence/absence to allow statistical analysis.
Osteoarthritis (OA) was considered present in cases with eburnation.
No distinction was made between active or <!-- ? --> lesions in cribra femora
and cribra orbitalia.
Data on chronic maxillary sinusitis were obtained from another study
[@casnaUrbanizationRespiratory2021].

[[ Description of method for sinusitis - Maia ]]
<!-- Description of lesions included in path conditions - Supp Material?

- OA = osteoarthritis
- IVDD = intervertebral disc disease
- TB = tuberculosis
- DISH = diffuse idiopathic skeletal hyperostosis
- VOP = vertebral osteophytosis
- SN = schmorl's nodes
- DDD = degenerative disc disease
- PNBF = periosteal new bone formation
- OD = osteochondritis dissecans
- CF = cribra femora
- CO = cribra orbitalia

Sinusitis as well
-->

@@

Caries ratios were calculated by dividing the number of lesions by the number
of teeth scored, resulting in a single caries ratio per individual. If the surface
where the lesion originated is not visible, i.e. if the lesion covered multiple
surfaces, this was scored as "crown".

Calculus indices were calculated according to Greene and colleagues [-@greeneQuantifyingCalculus2005].
Calculus was scored with a four-stage scoring system (0-3) to score absent, slight,
moderate, and heavy calculus deposits [@brothwellDiggingBones1981]
on the lingual, buccal (labial), and interproximal surfaces of each tooth. Only
one score was used for the combined interproximal surfaces, resulting in three
scores per tooth (when surfaces are intact), and four calculus indices per individual;
upper anterior (UA), upper posterior (UP), lower anterior (LA), lower posterior (LP).
Each index was calculated by dividing
the sum of calculus scores for each surface by the total number of surfaces scored
in each quadrant. If a tooth could not be scored on all three surfaces, the tooth
was not included [@greeneQuantifyingCalculus2005].

>>>>

New dental calculus samples were taken under sterile conditions in a positive
pressure laminar flow hood in a dedicated dental calculus lab at Leiden University.
The surface of the tooth was lightly brushed with a sterile, disposable toothbrush
to get rid of surface contaminants. A sterile dental curette was then used to
scrape calculus from the tooth onto weighing paper, which was transferred to 1.5 ml
Eppendorf tubes.
Where possible, previously sampled material was used to prevent repeated sampling
of individuals. The sampling methods can be found in the published material
[@ziesemer16SChallenges2015; @ziesemerGenomeCalculus2018]. 
The samples were sent to the Department of Forensic Medicine at Aarhus University
for ultra-high-performance liquid chromatography-tandem mass spectrometry method
using pneumatically assisted electrospray ionisation (UHPLC-ESI-MS/MS).

The UHPLC-ESI-MS/MS method was validated in a separate study on cadavers received
for forensic autopsy and toxicological analysis. Results from dental calculus were
validated against compounds detected in whole blood samples from the same
individuals.  
In the original method, samples were washed three times to remove residual
substances from the surface of the calculus that originated from oral fluids,
and only extract substances from the calculus. In our samples the washes served
to remove potential contaminants from the burial environment and post-excavation
handling.  
The list of targeted compounds included both naturally occurring compounds known
to have been used in the past, as well as synthetic modern drugs that did not exist
at the time (e.g. Fentanyl, MDMA, Amphetamine). These were part of the toxicology
screening for the original method [@sorensenDrugsCalculus2021],
developed on cadavers. In our study they serve as an authentication step, as their
presence in archaeological samples could only be the result of contamination.

Briefly, dental calculus was treated with citric acid and the dissolution extracts
were cleaned using weak and strong polymeric cation-exchange sorbents. Samples
were washed with 0.5 mL MeOH for 10 seconds. Samples were weighed before and after
each wash. Samples were air-dried for 24 hours at room temperature after each wash.
Extracts from each wash were analysed by injecting 5 mL volume into the column
of an Exion UHPLC system that consisted of two Exion AD pumps, an Exion AD
multiplate autosampler set at 10 $\pm$ 2 &deg;C and an Exion AC column oven set
at 40 $\pm$ 2 &deg;C (Sciex, Ontario, Canada).
Separation was performed using a Raptor Biphenyl UHPLC column (2.7 mm, 2.1 mm I.D.
$\times$100 mm) (Restek, Bellefonte, PA). The mass spectrometer was a Sciex QTRAP
6500+ with a TurboIonSpray probe for electrospray ionisation.  
The remaining calculus was dissolved using lysing tube beads in 800 $\mu$L of 0.5
$\small{M}$ citric acid (CA) and 50 $\mu$L stable isotope-labelled analogue used
as internal standards (SIL-IS) solution for 1 h at ambient
temperature with gentle shaking. The suspension was then mixed with 800 $\mu$L
MeOH and centrifuged at 10,000 $\times$ g for 5 mins, and analysed by the same
method as the wash extracts.
Data analysis was performed using Analyst 1.7 and MultiQuant 3.0.3 (Sciex).
Raw quantities of compounds are presented in ng and concentrations as ng / mg.  
For more details about the method and validation, see the original study by
Sørensen and colleagues [-@sorensenDrugsCalculus2021].

Of the `r nrow(demography)` individuals that were analysed,
`r nrow(filter(metadata, replicated == TRUE))` individuals were subsequently
replicated using leftover calculus from the original DNA study.
<!--in the same lab? different lab? -->

@@

<!-- methods for age, sex, and stature estimation -->
Demographic and pathological analyses were conducted in the Laboratory for Human
Osteoarchaeology at Leiden University. Sex was estimated using cranial
and pelvic morphological traits [@Standards1994]. Age-at-death was estimated using
dental wear,
auricular and pubic surface appearance, cranial suture closure, and epiphyseal fusion 
[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985], and divided into the following categories:
early young adult (EYA, 18-24), late young adult (LYA, 25-34), middle adult (MA, 
35-49), old adult (Old, 50+).

@@

Pathological conditions and lesions that occur frequently in the population were
included in the analysis. Data were
dichotomised to presence/absence to allow statistical analysis.
Osteoarthritis (OA) was considered present in cases where eburnation was visible
on one or more joint surface.
Vertebral osteophytosis is identified by marginal lipping and/or osteophyte
formation on the margin of the superior and inferior surfaces of the vertebral
body.
Cribra orbitalia (CO) was diagnosed based on presence of pitting on the superior
surface of the orbit. No distinction was made between active or <!-- ? --> lesions.
Degenerative disc disease (DDD), or spondylosis, is identified as a large diffuse
depression of the
superior and/or inferior surfaces of the vertebral body [@rogersPalaeopathologyJoint2000].
Schmorl's nodes (SN) are identified as any cortical depressions on the surface of
the vertebral body.
Data on chronic maxillary sinusitis from @casnaUrbanizationRespiratory2021 were
included in this study to assess the relationship between upper respiratory
diseases with environmental factors (i.e. tobacco smoke, caffeine consumption).
Lesions associated with chronic maxillary sinusitis (CMS) as defined by
@boocockMaxillarySinusitis1995 were
recorded for each individual and classified as "pitting", "spicule-type bone
formation", "remodeled spicules", or "white pitted bone". CMS was scored as absent
when the sinus presented smooth surfaces with little or no associated pitting.

@@

Caries ratios were calculated by dividing the number of lesions by the number
of teeth scored, resulting in a single caries ratio per individual. If the surface
where the lesion originated is not visible, i.e. if the lesion covered multiple
surfaces, this was scored as "crown".
Calculus indices were calculated according to Greene and colleagues [-@greeneQuantifyingCalculus2005].
Calculus was scored with a four-stage scoring system (0-3) to score absent, slight,
moderate, and heavy calculus deposits [@brothwellDiggingBones1981]
on the lingual, buccal (labial), and interproximal surfaces of each tooth. Only
one score was used for the combined interproximal surfaces, resulting in three
scores per tooth (when surfaces are intact), and four calculus indices per individual;
upper anterior (UA), upper posterior (UP), lower anterior (LA), lower posterior (LP).
Each index was calculated by dividing
the sum of calculus scores for each surface by the total number of surfaces scored
in each quadrant. If a tooth could not be scored on all three surfaces, the tooth
was not included [@greeneQuantifyingCalculus2005].
Periodontitis was scored on a visual four-stage (0-3) scoring system according to
distance from CEJ of each tooth to alveolar bone [@maatManualPhysical2005].
